A Method for Analysing Heterogeneous Data with 2D Detection Function Distance Sampling

### 2018-04-24 Draft.tex
4/24

@@

@@

@@

\subsection{Modification of plot functions}
The LT2D plot functions required slight modifications in order to be able to handle models with covariates included. Due to the infinite combination of choices of covariate values when these covariates are continuous (and growing according the the factorial of the number of levels for each of the factor level covariates), it was decided that producing plots for all covariate levels or for a fixed proportion of them was not possible. The plotting function therefore allows the user to select the integer number which represents the row of the original R data.frame which contains the detection who's covariate values are to be chosen as the values to be held fixed whilst plotting the various curves the plotting functions produce.  The option to overlay plots for a selection of these, or to manually specify a single set of covariate values which are to be held fixed would be desirable as the LT2D software develops further.

\section{Usage of the new software}
We now provide a selection of illustrative uses of the new LT2D software through the use of simulated data. All the following code is in R, with the LT2D package installed as the version included in the appendix of this project. The detection functions which constitute equations (4) through (8) inclusive are named h1, ip1, 1p2, ep1 and ep2 respectively within the LT2D package. The perpendicular density functions given by equations (12) through (16) inclusive are called pi.const, pi.hnorm, pi.norm, pi.chnorm and pi.chnorm.i respectively. We will begin our example by simulating a collection of detections from a population of size $N=500$, with an h1 detection function and a normal perpendicular density.

@@

# set perp trunc, forward trunc
# total transect length and survey
# area:
w = 0.03 ; ystart = 0.05
L = 10 ; A = 2*w*L

# set value of 'true' parameters
# for simulated data:
b <- c(-4.04,0.79)
logphi <- c(0.02,-4.42)

# produce simulated data:
set.seed(3)
simDat = simXY(500, 'pi.norm',
               logphi, 'h1', 
               b, w, 
               ystart)$locs
\end{lstlisting}
\endgroup

@@

\begin{center}
\tabcolsep=0.07cm
 \begin{tabular}{||c c c c c c c c||} 
 \hline
 x & y & stratum & transect & L & area & object & size\\ [0.5ex] 
 \hline\hline
 0.0050 & 0.0327 & 1 & 1 & 10 & 0.6 & 1 & 1 \\ 
 \hline
 0.0242 & 0.0322 & 1 & 1 & 10 & 0.6 & 2 & 1 \\
 \hline
 0.0115 & 0.0079 & 1 & 1 & 10 & 0.6 & 3 & 1 \\
 \hline
 0.0098 & 0.0305 & 1 & 1 & 10 & 0.6 & 4 & 1 \\
 \hline
 0.0180 & 0.0186 & 1 & 1 & 10 & 0.6 & 5 & 1 \\
 \hline
\end{tabular}
\end{center}

@@

\begingroup
\small
\begin{lstlisting}
# fit an LT2D model
fit <- LT2D.fit(DataFrameInput = sim.df,
                hr = 'h1',
                # start values for b:
                b = c(-3.72,0.74),
                ystart = ystart,
                pi.x = 'pi.norm',
                # start values for logphi:
                logphi = c(0.02, -4.42),
                w = w,
                hessian = TRUE)
\end{lstlisting}
\endgroup

\begin{figure}[H]
\center
\includegraphics[scale=0.85]{fig7}
\caption{Fitted curves for simulated data in the $x$ direction}
\end{figure}

Calls to the $gof.LT2D$, $LT2D.bootstrap$ and $plot$ functions produce goodness of fit vales for us, calculate a confidence interval on abundance using our chosen number of re-samples and plot the fitted curves. We can also extract the AIC from the fitted model object:

\begingroup
\small
\begin{lstlisting}
gof.LT2D(fit)
LT2D.bootstrap(fit,999)$ci
plot(fit)
fit$fit$AIC
\end{lstlisting}
\endgroup

Which gives us the GoF values:
\begin{center}
 \begin{tabular}{|c c c|} 
 \hline
 \  & KS & CvM \\ [0.5ex] 
 \hline
 X & 0.7130 & 0.7933 \\ 
 \hline
 Y & 0.9429 & 0.9300  \\
 \hline
\end{tabular}
\end{center}

Figures 7 and 8 display the fitted curves the simulated data. The bootstrap results in a 95\% confidence interval of $(490,.88,514.20)$ with 999 re-samples, from $n=448$ detections, with our fitted model producing an estimate of $\hat{N}=503$.  We add in some simulated covariate values to our data frame. We consider a continuous covariate $altitude.sim$ and a factor covariate $observer.id$, it should be noted that we are not further changing the simulated data other than by adding artifical columns into the data.frame that contains them, and so models including these fake covariates are expected to fit poorly. Their inclusion is there solely to illustrate how to specify covariates as part of a model within the LT2D software. The formulas argument to the $LT2D.fit$ function must be a list of R objects with class 'formula'. These are all of the form described in section 3.1. For each formula which is included within the model in this way, a vector of initial guesses must be passed to LT2D.fit, with the name ipars, xpars or ypars, depending on the formula specified. It is possible to specify multiple formulas at the same time, if the choice of detection function allows it (the software returns a warning and sometimes also an error if invalid combinations are selected by the user).

\begingroup
\footnotesize
\begin{lstlisting}
set.seed(3)

n <- length(simDat$x)

@@

iform <- formula(i~sim.df$observer.id)
xform <- formula(x~sim.df$altitude)
yform <- formula(y~sim.df$altitude)
# covariates in intercept only
fit2 <- LT2D.fit(DataFrameInput = sim.df,
                 hr = 'h1',
                 # start values for b:
                 b = c(-3.72,0.74),
                 ystart = ystart,
                 pi.x = 'pi.norm',
                 # start values for logphi:
                 logphi = c(0.02, -4.42),
                 w = w,
                 hessian = TRUE,
                 formulas = list(iform),
                 ipars=rep(0,2))

@@

# covariates in all of the h1 slots
fit4 <- LT2D.fit(DataFrameInput = sim.df,
                 hr = 'h1',
                 # start values for b:
                 b = c(-3.72,0.74),
                 ystart = ystart,
                 pi.x = 'pi.norm',
                 # start values for logphi:
                 logphi = c(0.02, -4.42),
                 w = w,
                 hessian = TRUE,
                 formulas=list(iform,
                 		xform,
                 		yform),
                 ipars=0
                 xpars=0,
                 ypars=0)
\end{lstlisting}
\endgroup

\section{Simulation studies}
Producing a simulation study to ascertain the new software's capacity to account for heterogeneous populations is made difficult by a number of factors. Firstly, as the \cite{Borchers} likelihood is complicated and requires multiple instances of numerical integration at every evaluation, model fitting and maximum likelihood estimation takes a significant amount of time. Models which have no covariate inclusion may take only a few seconds to be fitted to a relatively small sample, however, the inclusion of covariates increases this computational cost quite heavily, increasing the quantity of numerical integration involved (by having to evaluate each point individually, due to its own distinct estimated values of the detection function parameters, based on its covariate values) as well as requiring more calls to the likelihood function in order to numerically estimate its maximum likelihood estimates. This causes models with even relatively simple covariate formulas to sometimes take upwards of 15 minutes to compute. Therefore producing a simulation study which includes a large number of fitted models with covariate formulas is understandably difficult.

\begin{figure}[H]
\center
\includegraphics[scale=0.85]{fig8}
\caption{Fitted curves for simulated data in the $y$ direction}
\end{figure}

@@

A solution to this problem would be to include an option for the software to automatically recognise poor convergence in maximum likelihood estimates, and to perform a new estimation of the parameters using the previous estimates as initial values. This implementation would be complicated due to the desired flexibility of the covariate inclusion, and so is deemed too substantial to include as part of this project. We are therefore limited to simulation studies with very few fitted models, so that we may manually ensure that convergence for each model is reasonable, and therefore representative of the software's performance in practice. 
\section{Going forwards}
\subsection{Additional desired features}
\subsection{A proposed approach for mixture model incorporation}
\section{Conclusion}
This project's primary aim has been the implementation of covariates within the 2-Dimensional detection functions of the methodology provided by \cite{Borchers}, and to provide this additional functionality within the LT2D package itself, to increase its scope and feature set. The work completed for this project has allowed the LT2D software  to provide flexible methods for the user to deal with data which are heterogeneous with respect to detection probability, with a simple user interface. The user is able to choose any combination of covariates, be they quantitative or qualitative, and is able to specify separate formulas for each direction of the detection function. Plotting and goodness of fit functions have been generalised to be able to handle models with covariates included, with the possibility of calling the plot function multiple times to produce figures of the fitted curves at different levels of the included covariates. Not only has the abundance estimation within the software been generalised to produce point estimates when covariates have been included within a user's selected model, but it has also been extended to produce a bootstrap confidence interval of abundance (with the flexibility for the user to modify both the expected coverage and the bootstrap's  number of re-samples).

The implementation of this covariate inclusion required intimate knowledge of the \cite{Borchers} paper and the theory which underpins its software, as well as the ability to work with its substantial and complicated pre-existing code base to include the desired additional functionality. A strong grasp of good programming practice, relevant computer theoretic notions and comfort with the R programming language were all required to make the desired changes in an computationally efficient way. The use of techniques and practices outlined in \cite{Hadley} allowed the inclusion of more advanced functionality to the user by making use of R features which go above the scope of undergraduate statistics content (such as object oriented programming, exception handling, etc).

Due to the feature set of the new software, its flexibility and easy of use, it is my belief that the objectives outlined by this project have been comfortably met. Further examples of the software's use and demonstrations of it's ability to handle general cases can be found in the appendix.

@@

Given that the LT2D source code is approximately 3500 lines of R code (after the completion of this project), and that the inclusions which constitute this project are over 1500 alone, GitHub links to the relevant files will be included to avoid bloating this document.

@@

An example of how to fit different covariate formulas to a real data set:\\

\subsection{LT2D code}

>>>>

\begin{figure}[H]
\center
\includegraphics[scale=0.57]{gofx}
\caption{A perpendicular distance qq plot produced by the $gof.LT2D$ function for a model with covariate inclusion}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[scale=0.57]{gofy}
\caption{A forward distance qq plot produced by the $gof.LT2D$ function for a model with covariate inclusion}
\end{figure}

@@

\begin{figure}[H]
\center
\includegraphics[scale=0.57]{boothist}
\caption{A bootstrap distribution of abundance estimates resulting from a call to $LT2D.bootstrap$}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.57]{covarplot1}
\caption{Fitted detection function for an LT2D model at a given covariate level, using the modified plot function}
\end{figure}

@@

\subsection{Modification of plot functions}
The LT2D plot functions required slight modifications in order to be able to handle models with covariates included. Due to the infinite combination of choices of covariate values when these covariates are continuous (and growing according to the factorial of the number of levels for each of the factor level covariates), it was decided that producing plots for all covariate levels or for a fixed proportion of them was not reasonable. The plotting function therefore allows the user to select the integer number which represents the row of the original R data.frame which contains the detection who's covariate values are to be chosen as the values to be held fixed whilst plotting the various curves the plotting functions produce.  The option to overlay plots for a selection of these, or to manually specify a single set of covariate values which are to be held fixed would be desirable as the LT2D software develops further.

Figures 10 and 11 display the plots with respect perpendicular distance produced by the LT2D software for the same model (fitted to the same data), with different covariate values.

\begin{figure}[H]
\center
\includegraphics[scale=0.57]{covarplot2}
\caption{Fitted detection function for an LT2D model at a given covariate level, using the modified plot function}
\end{figure}

\section{Usage of the new software}
We now provide a selection of illustrative uses of the new LT2D software through the use of simulated data. All the following code is in R, with the LT2D package installed as the version included in the appendix of this project. The detection functions which constitute equations (4) through (8) inclusive are named h1, ip1, 1p2, ep1 and ep2 respectively within the LT2D package. The perpendicular density functions given by equations (12) through (16) inclusive are called pi.const, pi.hnorm, pi.norm, pi.chnorm and pi.chnorm.i respectively. We will begin our example by simulating a collection of detections from a population of size $N=50$, with an h1 detection function and a normal perpendicular density. The $b$ parameters in the code refer to detection function parameters, whereas $logphi$ refers to parameters of the perpendicular density function.

@@

# set perp trunc, forward trunc
# total transect length and survey
# area:
w = 0.15 ; ystart = 0.55
L = 10 ; A = 2*w*L

# set value of 'true' parameters
# for simulated data:
b=c(-7.3287948, 0.9945317)
logphi <- c(0.02,-4.42)

# produce simulated data:
set.seed(3)
simDat = simXY(50, 'pi.norm',
               logphi, 'h1', 
               b, w, 
               ystart)$locs
\end{lstlisting}
\endgroup

@@

\begin{center}
\tabcolsep=0.07cm
 \begin{tabular}{||c c c c c c c c||} 
 \hline
 x & y & stratum & transect & L & area & object & size\\ [0.5ex] 
 \hline\hline
 0.0174 & 0.0162 & 1 & 1 & 10 & 3 & 1 & 1 \\ 
 \hline
 0.0054 & 0.0223 & 1 & 1 & 10 & 3 & 2 & 1 \\
 \hline
 0.0156 & 0.0178 & 1 & 1 & 10 & 3 & 3 & 1 \\
 \hline
 0.0108 & 0.0303 & 1 & 1 & 10 & 3 & 4 & 1 \\
 \hline
 0.0277 & 0.0458 & 1 & 1 & 10 & 3 & 5 & 1 \\
 \hline
\end{tabular}
\end{center}

@@

\begingroup
\small
\begin{lstlisting}
# fit an LT2D model
fit <- LT2D.fit(DataFrameInput = sim.df,
                hr = 'h1',
                # start values for b:
                b = b,
                ystart = ystart,
                pi.x = 'pi.norm',
                # start values for logphi:
                logphi = logphi,
                w = w,
                hessian = TRUE)

gof.LT2D(fit)
\end{lstlisting}
\endgroup

The $gof.LT2D$ function gives us goodness of fit values for the fitted model in the perpendicular and forward directions. For the above model, the output of the $gof.LT2D function$ is:
\begin{center}
 \begin{tabular}{|c c c|} 
 \hline
 \  & KS & CvM \\ [0.5ex] 
 \hline
 X & 0.8810348 & 0.7445199 \\ 
 \hline
 Y & 0.9999873 & 0.9830506  \\
 \hline
\end{tabular}
\end{center}

We add in some simulated covariate values to our data frame. We consider a continuous covariate $altitude.sim$ and a factor covariate $observer.id$, it should be noted that we are not further changing the simulated data other than by adding artificial columns into the data.frame that contains them. The inclusion of these simulated covariates is solely to illustrate how to specify covariates as part of a model within the LT2D software. The formulas argument to the $LT2D.fit$ function must be a list of R objects with class 'formula'. These are all of the form described in section 3.1. For each formula which is included within the model in this way, a vector of initial guesses must be passed to LT2D.fit, with the name ipars, xpars or ypars, depending on the formula specified. If a user has selected a complicated formula, and is unsure on how many extra starting parameters must be supplied, they can use the $ParamNumRequired$ function to obtain an answer. It is possible to specify multiple formulas at the same time, if the choice of detection function allows it (the software returns a warning and sometimes also an error if invalid combinations are selected by the user). Once the model is fitted, we can also produce plots by specifying the row number of the input data frame which contains the covariate values of interest, as well as call the $LT2D.bootstrap$ function to produce a confidence interval with an expected coverage probability and chosen re-sample number.
This gives us a 95\% confidence interval of (25,53) for model $fit4$, as well as figures 12 through 15.\\

\begin{figure}[H]
\center
\includegraphics[scale=0.5]{fig13}
\caption{Fitted forward detection function for the first observation of $fit4$}
\end{figure}

The relevant code is found below:
\begingroup
\scriptsize
\begin{lstlisting}
set.seed(3)
n <- length(simDat$x)

@@

iform <- formula(i~sim.df$observer.id)
xform <- formula(x~sim.df$altitude)
yform <- formula(y~sim.df$altitude)
ParamNumRequired(sim.df, iform)

# covariates in intercept only
fit2 <- LT2D.fit(DataFrameInput = sim.df,
                 hr = 'h1',
                 # start values for b:
                 b = c(-3.72,0.74),
                 ystart = ystart,
                 pi.x = 'pi.norm',
                 # start values for logphi:
                 logphi = c(0.02, -4.42),
                 w = w,
                 hessian = TRUE,
                 formulas = list(iform),
                 ipars=rep(0,2))

@@

# covariates in all of the h1 slots
fit4 <- LT2D.fit(DataFrameInput = sim.df,
                 hr = 'h1',
                 # start values for b:
                 b = c(-3.72,0.74),
                 ystart = ystart,
                 pi.x = 'pi.norm',
                 # start values for logphi:
                 logphi = c(0.02, -4.42),
                 w = w,
                 hessian = TRUE,
                 formulas=list(iform,
                  	       xform,
                 	       yform),
                 ipars=rep(0,2),
                 xpars=0,
                 ypars=0)

plot(fit4, covar.row=1,smooth.fy=T)
plot(fit4, covar.row=2,smooth.fy=T)
boot <- LT2D.bootstrap(fit4,r=999,alpha=0.05)
boot$ci
\end{lstlisting}
\endgroup

\begin{figure}[H]
\center
\includegraphics[scale=0.53]{fig12}
\caption{Fitted perpendicular curves for the first observation of $fit4$}
\end{figure}

\section{Simulation studies}
Producing a simulation study to ascertain the new software's capacity to account for heterogeneous populations is made difficult by a number of factors. Firstly, as the \cite{Borchers} likelihood is complicated and requires multiple instances of numerical integration at every evaluation, model fitting and maximum likelihood estimation takes a significant amount of time. Models which have no covariate inclusion may take only a few seconds to be fitted to a relatively small sample, however, the inclusion of covariates increases this computational cost quite heavily, increasing the quantity of numerical integration involved (by having to evaluate each point individually, due to its own distinct estimated values of the detection function parameters, based on its covariate values) as well as requiring more calls to the likelihood function in order to numerically estimate its maximum likelihood estimates. This causes models with even relatively simple covariate formulas to sometimes take upwards of 15 minutes to compute. Therefore producing a simulation study which includes a large number of fitted models with covariate formulas is understandably difficult.\\\\

\begin{figure}[H]
\center
\includegraphics[scale=0.5]{fig14}
\caption{Fitted perpendicular curves for the second observation of $fit4$}
\end{figure}

\begin{figure}[H]
\center
\includegraphics[scale=0.5]{fig15}
\caption{Fitted forward detection function for the second observation of $fit4$}
\end{figure}

@@

A solution to this problem would be to include an option for the software to automatically recognise poor convergence in maximum likelihood estimates, and to perform a new estimation of the parameters using the previous estimates as initial values. This implementation would be complicated due to the desired flexibility of the covariate inclusion, and so is deemed too substantial to include as part of this project. We are therefore limited to simulation studies with very few fitted models, so that we may manually ensure that convergence for each model is reasonable, and therefore representative of the software's performance in practice.

The final issue with producing simulation studies to compare the bias between covariate-including and covariate-excluding models lies within the capacities of the LT2D package's simulation functions. These functions currently have no way of generating heterogeneous data with a single call to the function, and no way of incorporating any covariate which is not a factor. The inclusion of this functionality would be extremely desirable, in order to be able to further verify the effectiveness of covariate inclusion within the LT2D package, however this would represent an extension which is above the scope of this project, and therefore must be left as the subject of further research.

We conduct our

\begin{center}
 \begin{tabular}{||c c c||} 
 \hline
 seed & covariate est & no-covariate est  \\ 
 \hline
 1 & 267.6 & 265.7 \\
 \hline
 2 & 197.1 & 195.7  \\
 \hline
 3 & 173.1 & 173.4  \\
 \hline
 4 & 199.3 & 201.4  \\
 \hline
 5 & 214 & 224.4 \\
 \hline
 6 & 210.8 & 209.4 \\
 \hline
 7 & 162.1 & 155.4 \\
 \hline
 8 & 190.5 & 188.1 \\
 \hline
 9 & 212.7 & 227 \\
 \hline
 10 & 156.4 & 157.2 \\
 \hline
 11 & 164.4 & 169.2 \\
 \hline
 12 & 165.6 & 168.9 \\
 \hline
 13 & 179.7 & 177.6 \\
 \hline
 14 & 197.9 & 198.8 \\
 \hline
 15 & 162.5 & 163.6 \\
 \hline
 mean & 190.25 & 191.72 \\
 \hline
\end{tabular}
\end{center}

\section{A proposed approach for mixture model incorporation}
\section{Conclusion}
This project's primary aim has been the implementation of covariates within the 2-Dimensional detection functions of the methodology provided by \cite{Borchers}, and to provide this additional functionality within the LT2D package itself, to increase its scope and feature set. The work completed for this project has allowed the LT2D software  to provide flexible methods for the user to deal with data which are heterogeneous with respect to detection probability, with a simple user interface. The user is able to choose any combination of covariates, be they quantitative or qualitative, and is able to specify separate formulas for each direction of the detection function. Plotting and goodness of fit functions have been generalised to be able to handle models with covariates included, with the possibility of calling the plot function multiple times to produce figures of the fitted curves at different levels of the included covariates. Not only has the abundance estimation within the software been generalised to produce point estimates when covariates have been included within a user's selected model, but it has also been extended to produce a bootstrap confidence interval of abundance (with the flexibility for the user to modify both the expected coverage and the bootstrap's  number of re-samples).

The implementation of this covariate inclusion required intimate knowledge of the \cite{Borchers} paper and the theory which underpins its software, as well as the ability to work with its substantial and complicated pre-existing code base to include the desired additional functionality. A strong grasp of good programming practice, relevant computer theoretic notions and comfort with the R programming language were all required to make the desired changes in a computationally efficient way. The use of techniques and practices outlined in \cite{Hadley} allowed the inclusion of more advanced functionality to the user by making use of R features which go above the scope of undergraduate statistics content (such as object oriented programming, exception handling, etc).

Due to the feature set of the new software, its flexibility and easy of use, it is my belief that the objectives outlined by this project have been comfortably met. Further examples of the software's use and demonstrations of its ability to handle general cases can be found in the appendix.

@@

Given that the LT2D source code is approximately 3500 lines of R code (after the completion of this project), and that the inclusions which constitute this project are over 1500 alone, GitHub links to the relevant files will be included, rather than raw code, to avoid bloating this document.\\

@@

An example of how to fit different covariate formulas to a real data set (from the file which was used to fit examples useful for debugging during the development phase of the project):\\
